---
title: "Tidyverse"
author: "Luiz Felipe & Francisco Palmero"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 1) Short review

-   Creating a chunk of code (ctrl + alt + i)
-   Loading packages (tidyverse and readxl)
-   Creating an object
-   Commenting in .Rmd: in-chunk or out-of-chunk
-   Running a chunk of code
-   View an object in R studio
-   Pipe operator

# 2) Load packages

```{r}
library(tidyverse)
library(readxl)
```

# 3) Import data

```{r}
# Load weather data
df_weather <- read.csv("../Data/nfix_weather.csv") %>% dplyr::select(-X) # Select (-X) Borra la primer columna que indica el número de fila
# Load biomass data
df_biomass <- read.csv("../Data/nfix_biomass.csv") %>% dplyr::select(-X) 

# Load yield data
df_yield <- read.csv("../Data/nfix_yield.csv") %>% select(-X)

str(df_weather)
str(df_biomass)
str(df_yield)

view(df_biomass)
```

# 4) Tidyverse

-   What's tidy data?

### SELECT

Helper functions:
- starts_with() Me muestra las columnas que empiecen con lo que le pida ".." y concatenados si son mas de una
- ends_with()   ""   ""     ""    ""    ""  terminan ""  ""  ""  "  ""
- contains()
- matches()
- num_range()
- grepl() Busca y combina patrones dentro de un elemento que difieren en el formato o carácter de resultados.
- between() Detecta los valores que estan en un rango especificado
- across() Me facilita realizar la misma transformación a múltiples columnas

```{r}
# Select specific columns from the data
df_weather %>% 
  select(season)

# Select columns whose names start with a given string
df_weather %>% 
  select(starts_with(c("PP","Tmean")))

# Select columns whose names end with a specific string.
df_weather %>% 
  select(ends_with("A"))

df_weather %>% 
  
```

### ARRANGE me ordena los datos de la columna que le indique de menor a mayor

```{r}
# Reorder the data frame rows according to the lowest-to-highest yield
df_yield %>% dplyr::arrange(Yield_mgha)

# Reorder the data frame rows according to the highest-to-lowest yield
df_yield %>% 
  arrange(desc(Yield_mgha))
```

### RENAME me permite cambiar el nombre de las columnas primero el nombre que quiero = la forma q estaba

```{r}
# Change column name
df_yield %>% 
  rename(yield = Yield_mgha)

# To upper or lower case (rename_with(tolower/touuper)) Pasa todas las columnas a lower case "minúscula"
df_yield %>% 
  rename_with(tolower)

df_yield %>%  #Pasa todas las columnas a upper case "mayúscula"
  rename_with(toupper)
```

### FILTER permite filtrar dentro de un objeto o df la columna que le indique

```{r}
# Filter rows
df_weather %>% filter(PP_A <= 20)

# Filter rows
df_weather %>% filter(Site == "IN1")

# Filter rows where Yield_mgha is between 5 and 7 tons/ha (between() or using &)
df_yield %>% filter(between(Yield_mgha, 5, 7))
df_yield %>% filter(Yield_mgha > 5 & Yield_mgha < 7)

# Filter KS1 sites (`%in%`)
df_weather %>% filter(Site %in% "KS1")

# Filter KS1 and KS2 sites (`%in%`)
df_weather %>% filter(Site %in% c("KS1","KS2"))

# Filter rows where Yield_mgha is not NA (!is.na())
df_yield %>% filter(is.na(Yield_mgha))

# Filter rows where Site contains "KS" (grepl) me filtra todos los sitios a excepción de los que tienen KS
df_weather %>% filter(!grepl("KS", Site))
```

### MUTATE me permite crear nuevas columnas en el data frame

```{r}
# Create a new column with total PP
df_weather %>% mutate(PPtotal = PP_A + PP_B + PP_C) #Crea la columna PPtotal con la suma que le pedí y por
#default lo manda al final del data drame

# Create a new column with total PP after a given column
df_weather %>% mutate(PPtotal = PP_A + PP_B + PP_C, .after = "season")
df_weather %>% mutate(PPtotal = PP_A + PP_B + PP_C, .before = "Dur_A") #de esta manera le indico donde quiero la nueva columna

# Change type of data from character to factor (`mutate(across(c()))`)
df_biomass %>% mutate(across(c(Site, Stage, Block, Treatment), as.factor))
```

### ROUND 

```{r}
# Round yield values
df_yield %>% mutate(across(Yield_mgha,round,digits=2))
```

### SUMMARISE me aporta datos resumen del data frame como media, mediana, desvio estandar

```{r}
# Create a summary data frame with mean yield (na.rm=T)
df_yield %>% summarise(mean = mean(Yield_mgha))
df_yield %>% summarise(sd = sd(Yield_mgha))
df_yield %>% summarise(median = median(Yield_mgha))

summary(df_yield)
```

### GROUP_BY agrupa una o mas variables

```{r}
# Group by Treatment and summarize mean yield
df_yield %>% 
  group_by(Treatment) %>% 
  summarise(mean = mean(Yield_mgha, na.rm = TRUE))

# Group by Site and Treatment and summarize mean yield
df_yield %>% 
  group_by(Treatment, Site) %>% 
  summarise(mean = mean(Yield_mgha, na.rm = TRUE))
```

### SEPARATE separa elementos que estan en la misma columna en nuevas columnas

```{r}
# Separate Site into Location and Year
df_yield %>% separate(Site, c("Location","Season"), remove = F, sep = 3)
df_yield %>% separate(Site, c("Location","Season"), remove = F, sep = "_")
```

### UNITE combina varias columnas a una sola especificada con el nombre

```{r}
# Create ID column (remove or not previous)
df_biomass %>% unite(ID, Site, Stage, Block, Treatment, sep = "_")
```

### DISTINCT reconoce cuantas cosas distintas hay en una columna

```{r}
# Get distinct Sites (distinct())
df_yield %>% distinct(Treatment)

# Get distinct combinations of Site and Treatment
df_yield %>% distinct(Site, Treatment)

df_yield %>% 
  group_by(Site, Treatment, Block) %>% 
  unique() %>% 
  summarise()
```

### COUNT

```{r}
# Count number of rows within each site (count())
df_yield %>% 
  count(Site)
```

### CASE_WHEN

```{r}
# Create a new column that categorizes yields into "low", "medium", and "high"
# Yield_mgha < 3.5 ~ "low"; Yield_mgha >= 3.5 & Yield_mgha < 5 ~ "medium"; Yield_mgha >= 5 ~ "high"
df_yield %>% 
  mutate(yield_class = case_when(Yield_mgha < 3.5 ~ "low",
                                 Yield_mgha >= 3.5 & Yield_mgha < 5 ~ "medium",
                                 Yield_mgha >= 5 ~ "high"))
df_yield %>% 
  mutate(yield_class = case_when(Yield_mgha < 3.5 ~ "low",
                                 Yield_mgha >= 3.5 & Yield_mgha < 5 ~ "medium",
                                 Yield_mgha >= 5 ~ "high"), .after = "Block")

# Change only "Check" to "Control" (T~)
df_biomass %>% 
  mutate(Treatment = case_when(Treatment == "Check" ~ "Control",
                               Treatment == "Unfertilized" ~ "Corn",
                               T ~ Treatment))

# Use base R to change the values of cells
df_yield[1, 2] <- "Cheeck"
df_yield[8, 2] <- "check"

df_yield %>% group_by(Treatment) %>% unique() %>% summarise()

# Solve the problem using %in%
df_yield <- df_yield %>%
  mutate(Treatment = case_when(
    Treatment %in% c("Cheeck", "check") ~ "Check", 
    TRUE ~ Treatment))
```

### PIPE

```{r}
# Chain multiple operations to process the data
df_yield %>%  
  filter(Treatment == "Check") %>% 
  group_by(Site) %>%
  mutate(mean_Yield_mgha = mean(Yield_mgha)) %>% 
  summarise(mean_Yield_mgha) %>% 
  unique() %>% 
  separate(Site, into = c("Location", "Season"), sep = "_") %>% 
  rename(meanY = mean_Yield_mgha)
```

### STR_REPLACE 

```{r}
# Replace a string in the Site column
df_yield %>% 
  mutate(across(Site, ~str_replace(., pattern = "_", replacement = ":")))

# Let's create a non-desired pattern. Let the error be an underscore in the Check treatment.
# Solve the error so that you have the true name back 
df_yield %>% 
  mutate(Treatment = case_when(Treatment == "Check" ~ "C_heck",
                                TRUE ~ Treatment)) %>% 
  mutate(across(Treatment,
                ~ str_replace(., pattern = "_",
                              replacement = "")))
```

### IS.NA and NA_IF

```{r}
# Example Dataset: df_yield
df_yield_NA <- tibble(
  Site = c("KS1", "KS3", "KS4"),
  Treatment = c("Check", "N", "S"),
  Yield = c(50, 70, 65))

# 1. Use `na_if()` to replace the value "Check" in the Treatment column with NA
df_yield_naif <- df_yield %>%
  mutate(Treatment = na_if(Treatment, "Check"))
 
print(df_yield_naif)

# 2. Use `is.na()` to create a new column that flags where Treatment is NA
df_yield_filtered <- df_yield_naif %>%
   mutate(Flag_NA = is.na(Treatment))

# 3. Use `is.na()` to filter out rows where the Treatment is NA
df_yield_filtered <- df_yield_naif %>%
   filter(!is.na(Treatment))
```

### JOIN FUNCTIONS

```{r}
# Sample data frames
df_X <- tibble(
  Site_Season = c("IA1_2021", "IA2_2021", "IA3_2022"),
  Yield = c(150, 200, 180))

df_Y <- tibble(
  Site_Season = c("IA1_2021", "IA2_2021", "IA4_2022"),
  Yield = c(50, 60, 55))

# INNER-JOIN: Includes common rows for x and y
df_X %>% 
  inner_join(df_Y, by = "Site_Season")

# LEFT_JOIN: Includes all rows from X and matched rows from Y
df_X %>% 
  left_join(df_Y, by = "Site_Season")

# RIGHT_JOIN: Includes all rows from Y and matched rows from X
df_X %>% 
  right_join(df_Y, by = "Site_Season")

# FULL_JOIN: Includes all rows from both X and Y
df_X %>% 
  full_join(df_Y, by = "Site_Season")
```

### BIND ROWS AND COLUMNS

### PIVOT_WIDER

```{r}
# Pivot wider: Convert long data back to wide format
df_yield %>% 
  pivot_wider(names_from = Treatment, values_from = Yield_mgha)
```

### PIVOT LONGER

```{r}
# Pivot longer: Convert wide data to long format
df_weather %>% 
  dplyr::select(Site, season, PP_A, PP_B, PP_C, PP_D) %>% 
  pivot_longer(cols = starts_with("PP"), names_to = "Period", values_to = "Precipitation") 
```

# 5) Final comments

-   Put all library() calls at the top of the script.
-   Organize your files into 'data', 'script/notebook'/'markdown', and 'output'.
-   Comment your code.
-   Object names: Don’t start with numbers & Keep them as short as understandable.
-   Each %>% should be followed by a new line.
-   Avoid long lines (start a new line after a comma).
-   Always use na.rm = TRUE in summarise() functions to handle NA values.
-   Use %>% for chaining functions to keep code clean and readable.
-   When joining datasets, ensure that the keys have consistent names and formats.
